FITS format for planetary surfaces: definitions, applications and best practices.

### 2018-02-27 ess_marmo.tex

\subsection{FITS Format}
FITS is an open digital standard, created in the late 1970's for data acquisition, transfer and
archiving of telescope data by astronomical observatories.
It had been used data exchange and archiving from several orbital telescopes.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
FITS data storage is compatible with the Planetary Data System (PDS)
archiving specifications so that FITS files can be embedded in PDS datasets.
Finally, it is supported by a large number of open libraries
and software tools.

@@

\begin{table}
\caption{Projection names and codes already available in FITS suitable for
planetary applications: the corresponding PDS4 terminology is
provided.}
\centering
\begin{tabular}{l l l}
\hline
FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
\hline
Zenithal Equidistant & ARC & Azimuthal Equidistant \\
Zenithal Perspective & AZP & General Vertical Near-sided Projection \\
Plate Carre & CAR & Equirectangular, Miller Cylindrical \\
Equidistant Conic & COD & Equidistant Conic \\
Conic Equal-Area & COE & Albers Conical Equal Area \\ 
Conic Ortomorphic & COO & Lambert Conformal Conic\\
Mercator & MER & Mercator, Oblique Mercator, Transverse Mercator \\
Polyconic & PCO & Polyconic \\
Sanson-Flamsteed & SFL & Sinusoidal \\
Ortographic & SIN & Orthographic \\
Stereographic & STG & Stereographic, Polar Stereographic \\
Gnomonic & TAN & Gnomonic \\
Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
\hline
\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}. We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator.}
\end{tabular}
\label{table:proj}
\end{table}

@@

\subsubsection{FITS WCS keywords and their recommended use in planetary surface context}
\label{sssec:wcskeys}
In this section WCS FITS keywords are inserted in the planetary surface context.
When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving and distribution
standardization purposes.
\begin{enumerate}
\item{RADESYS.
Name of the inertial reference frame with respect to which the coordinate
system is defined (defaulted to \texttt{`ICRS'})}
\item{EQUINOX.
The value field contains the equinox in years for the celestial
coordinate system in which positions are expressed (defaulted to \texttt{`J2000'})}
\item{WCSAXES.
Number of axes in the WCS description.
The value of WCSAXES may exceed the number of raster pixel axes. WCSAXES is, e.g.,
generally set to $2$ in two-dimensional rasters, it could be set to $3$ if an
additional time dependency is added to the coordinate description.}
\item{WCSNAME.
Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic,
non-rotating, inertial).
EPN-TAP provides a \texttt{spatial\_coordinate\_description} parameter.}
\item{CRPIXn.
Location of a reference point along axis $n$, in units of the axis index.
The reference point value do not need to lie within the actual data array.}
\item{CTYPEn.
Name of the coordinate represented by axis $n$, in ``4--3'' form: the first four characters
specify the coordinate type, the fifth character is a `-' and the
remaining three characters specify an algorithm code for computing the world coordinate
value (\ref{table:proj}).
For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG, where x or yz
correspond to the target body. We propose to standardize those codes as listed in Table
\ref{table:ctype}.
For projected coordinates in meters we propose a similar scheme: xPX as Projected X, xPY as
Projected Y, or yzPX and yzPY.
Alternative representation of WCS are already possible in FITS, to describe images in degrees
and in meters \citep{greisencalaII}.
The name of the alternative axis will be specified using the alternative CNAMEn keyword.}
\item{CRVALn.
Value of the coordinate specified by the CTYPEn keyword at the reference point CRPIXn.}
\item{CDi\_j.
The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the intermediate
world coordinates.}
\item{PCi\_j.
The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate pixel
coordinates (to be used with CDELT keywords).}
\item{CDELTn.
The coordinate increment along axis $n$ (scaling component of the linear transformation
matrix PCi\_j).
In PDS4 they correspond to the \textit{pixel\_resolution\_x} and \texttt{pixel\_resolution\_y}
attributes.
In EPN-TAP they are connected to the \texttt{spatial\_resolution} parameter.}
\item{CUNITn.
Units of the coordinate system , not mandatory, defaulted to deg for angular coordinates
and to meters (m) for linear ones.}
\item{CNAMEn.
The description of a particular coordinate for an axis in a particular WCS (up to 68
characters), while the WCSNAMEa keyword names the particular WCS as a whole.
Its default value will be blank.}
\end{enumerate}

@@

\item{\texttt{BLANK}.
This keyword must be used only with integer data.
It contains an integer that specifies the value that is used to
represent pixels that have an undefined physical value.
On floating point data the IEEE NaN must be used to represent undefined values.  
\texttt{PDS4 Special\_Constants?}}

@@

\subsection{MATISSE}
MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool for the
Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool designed by the Space
Science Data Center of the Italian Space Agency (SSDC-ASI) to access, visualize and analyse data from
solar system exploration missions.
Currently, it provides access to public data from 5 different targets (i.e., Mercury, Venus, Mars,
dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local storage or by querying
remote dataset, such as VESPA, Planet Server and NASA JPL (by means of the ODE REST protocol) and
can be integrated with other datasets thanks to its modular structure.
Apart from its peculiar 3D visualization, it is worthy to note that one of its 2D outputs is
formatted as FITS, projected by interpolating the original observations to a grid regularly spaced
in planetocentric coordinates (i.e., latitude and longitude).
We used the planetary FITS standard described in this work in order to make the output file
compliant with most FITS readers, especially DS9 and JS9.
The files generated by MATISSE could be also opened with standard GIS software (e.g., QGIS, ArcGIS) 
after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate projection.
Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be online by the end
of 2018) we are planning to make use of JS9 as the standard reader for the 2D online output of the
tool, thus allowing users to perform advanced analysis in both 2D and 3D directly inside a web-based
environment.

\subsection{Aladin}
Aladin is both a sky atlas and a portal to access data -- mainly images and
catalogues -- available through the VO.
It is able to load FITS files and to display images taking into account various
WCS projections.
Like DS9, Aladin is popular and broadly used in the part of the astronomical
community manipulating images.
So far it is also the reference generator and visualizer of Hierarchical
Progressive Survey (HiPS).
HiPS\footnote{http://www.ivoa.net/documents/HiPS/} is a recommendation endorsed
by the International Virtual Observatory Alliance (IVOA).
It describes how to access to sky survey data stored at various spatial
resolutions, offering a progressive view of possibly very large datasets.
The purpose is to allows users to explore all-sky surveys without having to
explicitly perform any query: the most important information is visible at low
spatial resolutions while finer details appear by zooming in.
In addition, an important property of HiPS is to preserve as much as possible
the scientific content of the original dataset (few spatial distortions,
flux conservation).

In the framework of the VESPA project, the idea of using Aladin to display
planetary data, especially planetary HiPS, has emerged.
As a general tool, not dedicated to planetology, Aladin does not support PDS
files.
HiPS are usually generated from a collection of WCS tagged FITS files.
%As an alternative, Aladin can use JPEG files together with additional metadata
%files.
%A metadata file -- one by JPEG file -- basically contains in a text format the
%WCS part of a FITS header.

Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
been updated to support several planetary conventions.
First, while the sky is displayed from the inside of the unit sphere, planetary
surfaces are observed -- and must be displayed -- from the exterior.
It means that the direction of the east on the horizontal axis has to be
inverted when displaying planetary data.
Then, planetary missions usually use different projections when providing
images of planets polar caps and equatorial region.
Often the former are in the stereographic (STG) projection while the later
is in the Plate Carr\'ee (CAR) projection. 
This specificity has been taken into account into the HiPS generator.
Also, in the future, Aladin will be able to recognize the FITS keywords defined
in the present paper (see Tab. \ref{table:proj} and \ref{table:ctype}).
For example, it will be able to take into account a planet's radii to compute
distances on the surface of the body.
Finally, among the WCS projections listed in Tab. \ref{table:proj}, several of them are
barely used in extrasolar astronomy and are not supported in Aladin yet.
An effort has been started to revamp and update the code of projections in
Aladin.
We expect all projections to be available in a forthcoming release.

>>>>

\subsection{FITS Format}
FITS is an open digital standard, created in the late 1970's for data acquisition,
transfer and archiving of telescope data by astronomical observatories.
It had been used for data exchange and archiving from several orbital telescopes.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard formats implemented in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
FITS data storage is compatible with the Planetary Data System (PDS)
archiving specifications so that FITS files can be embedded in PDS datasets.
Finally, it is supported by a large number of open libraries
and software tools.

@@

\begin{table}
\caption{Projection names and codes already available in FITS suitable for
planetary applications: the corresponding PDS4 terminology is
provided.}
\centering
\begin{tabular}{l l l}
\hline
FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
\hline
Zenithal Equidistant & ARC & Azimuthal Equidistant \\
Zenithal Perspective & AZP & General Vertical Near-sided Projection \\
Plate Carre & CAR & Equirectangular, Miller Cylindrical \\
Equidistant Conic & COD & Equidistant Conic \\
Conic Equal-Area & COE & Albers Conical Equal Area \\ 
Conic Ortomorphic & COO & Lambert Conformal Conic\\
Mercator & MER & Mercator, Oblique Mercator, Transverse Mercator \\
Polyconic & PCO & Polyconic \\
Sanson-Flamsteed & SFL & Sinusoidal \\
Ortographic & SIN & Orthographic \\
Stereographic & STG & Stereographic, Polar Stereographic \\
Gnomonic & TAN & Gnomonic \\
Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
\hline
\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}. We cannot find correspondences in the WCS scheme for PDS4 Robinson,} \\
\multicolumn{3}{l}{Van Den Grinten and Space Oblique Mercator: those three projection are not used for archival or analysis.}
\end{tabular}
\label{table:proj}
\end{table}

@@

\subsubsection{FITS WCS keywords and their recommended use in planetary surface context}
\label{sssec:wcskeys}
In this section WCS FITS keywords are inserted in the planetary surface context.
When possible, corresponding PDS4 and EPN-TAP definitions are added for archiving
and distribution standardization purposes.
\begin{enumerate}
\item{RADESYS.
Name of the inertial reference frame with respect to which the coordinate
system is defined (defaulted to \texttt{`ICRS'})}
\item{EQUINOX.
The value field contains the equinox in years for the celestial
coordinate system in which positions are expressed (defaulted to \texttt{`J2000'})}
\item{WCSAXES.
Number of axes in the WCS description.
The value of WCSAXES may exceed the number of raster pixel axes. WCSAXES is, e.g.,
generally set to $2$ in two-dimensional rasters, it could be set to $3$ if an
additional time dependency is added to the coordinate description.}
\item{WCSNAME.
Name of world coordinate descriptions (e.g. body-fixed rotating -ocentric -ographic,
non-rotating, inertial).
EPN-TAP provides a \texttt{spatial\_coordinate\_description} parameter.}
\item{CRPIXn.
Location of a reference point along axis $n$, in units of the axis index.
The reference point value do not need to lie within the actual data array.}
\item{CTYPEn.
Name of the coordinate represented by axis $n$, in ``4--3'' form: the first four
characters specify the coordinate type, the fifth character is a `-' and the
remaining three characters specify an algorithm code for computing the world coordinate
value (\ref{table:proj}).
For angular coordinates \citet{calagreisenI} provides the form xLN xLG or yzLN yzLG,
where x or yz correspond to the target body.
We propose to standardize those codes as listed in Table \ref{table:ctype}.
For projected coordinates in meters we propose a similar scheme: xPX as Projected X,
xPY as Projected Y, or yzPX and yzPY.
Alternative representation of WCS are already possible in FITS, to describe images in
degrees and in meters \citep{greisencalaII}.
The name of the alternative axis will be specified using the alternative CNAMEn
keyword.}
\item{CRVALn.
Value of the coordinate specified by the CTYPEn keyword at the reference point CRPIXn.}
\item{CDi\_j.
The $ij$ linear transformation matrix (rotation, skewness, scaling) defining the
intermediate world coordinates.}
\item{PCi\_j.
The $ij$ linear transformation matrix (rotation, skewness) defining the intermediate
pixel coordinates (to be used with CDELT keywords).}
\item{CDELTn.
The coordinate increment along axis $n$ (scaling component of the linear transformation
matrix PCi\_j).
In PDS4 they correspond to the \textit{pixel\_resolution\_x} and
\texttt{pixel\_resolution\_y} attributes.
In EPN-TAP they are connected to the \texttt{spatial\_resolution} parameter.}
\item{CUNITn.
Units of the coordinate system, not mandatory, defaulted to degrees for angular
coordinates and to meters (m) for linear ones.}
\item{CNAMEn.
The description of a particular coordinate for an axis in a particular WCS (up to 68
characters), while the WCSNAMEa keyword names the particular WCS as a whole.
Its default value will be blank.}
\end{enumerate}

@@

\item{\texttt{BLANK}.
This keyword must be used only with integer data.
It contains an integer that specifies the value that is used to
represent pixels that have an undefined physical value.
On floating point data the IEEE NaN must be used to represent undefined values.  
It is comparable with the \texttt{invalid\_constant} or \texttt{missing\_constant}
in PDS4 \texttt{Special Constants} class.}

@@

\subsection{MATISSE}
MATISSE\footnote{https://tools.asdc.asi.it/matisse.jsp} (Multi-purpose Advanced Tool
for the Instruments for the Solar System Exploration) \citep{ZINZI2016} is the webtool
designed by the Space Science Data Center of the Italian Space Agency (SSDC-ASI) to
access, visualize and analyze data from solar system exploration missions.
Currently, it provides access to public data from five different targets (i.e., Mercury,
Venus, Mars, dwarf planet Ceres and asteroid Vesta) using both data stored on SSDC local
storage or by querying remote dataset, such as VESPA, Planet Server and NASA JPL and
can be integrated with other datasets thanks to its modular structure.
One of th MATISSE 2D outputs is formatted as FITS, projected by interpolating the
original observations to a grid regularly spaced
in planetocentric coordinates (i.e., latitude and longitude).
We used the planetary FITS standard described in this work in order to make the output
file compliant with most FITS readers, especially DS9 and JS9.
The files generated by MATISSE could be also opened with standard GIS software (e.g.,
QGIS, ArcGIS) 
after being provided as input of the fits2vrt Python library\footnote{https://github.com/epn-vespa/fits2vrt},
whose output is a VRT virtual GDAL header pointing to the FITS file with the appropriate
projection.
Thanks to this approach, in the next major release of MATISSE (v2.0, planned to be
online by the end of 2018) we are planning to make use of JS9 as the standard reader for
the 2D online output of the tool, thus allowing users to perform advanced analysis in
both 2D and 3D directly inside a web-based environment.

\subsection{Aladin}
Aladin is both a sky atlas and a portal to access data -- mainly images and
catalogues -- available through the VO.
It is able to load FITS files and to display images taking into account various
WCS projections.
Like DS9, Aladin is popular and broadly used in the part of the astronomical
community manipulating images.
So far it is also the reference generator and visualizer of Hierarchical
Progressive Survey (HiPS).
HiPS\footnote{http://www.ivoa.net/documents/HiPS/} is a recommendation endorsed
by the International Virtual Observatory Alliance (IVOA).
It describes how to access to sky survey data stored at various spatial
resolutions, offering a progressive view of possibly very large datasets.
%The purpose is to allows users to explore all-sky surveys without having to
%explicitly perform any query: the most important information is visible at low
%spatial resolutions while finer details appear by zooming in.
%In the framework of the VESPA project, the idea of using Aladin to display
%planetary data, especially planetary HiPS, has emerged.
HiPS are usually generated from a collection of WCS tagged FITS files.
%As an alternative, Aladin can use JPEG files together with additional metadata
%files.
%A metadata file -- one by JPEG file -- basically contains in a text format the
%WCS part of a FITS header.
Aladin\footnote{http://aladin.u-strasbg.fr/} \citep{aladin} and its HiPS generator have
been updated to support planetary conventions.
%Then, planetary missions usually use different projections when providing
%images of planets polar caps and equatorial region.
%Often the former are in the stereographic (STG) projection while the later
%is in the Plate Carr\'ee (CAR) projection. 
%This specificity has been taken into account into the HiPS generator.
In the future, Aladin will be able to recognize the FITS keywords defined
in the present paper (see Tab. \ref{table:proj} and \ref{table:ctype}).
For example, it will be able to take into account a planet's radii to compute
distances on the surface of the body.
Finally, among the WCS projections listed in Tab. \ref{table:proj}, several of them are
barely used in extrasolar astronomy and are not supported in Aladin yet.
An effort has been started to revamp and update the code of projections in
Aladin.
We expect all projections to be available in a forthcoming release.
